\pkg{shorts}: An R Package for Modeling Short Sprints

### 2020-10-21 paper.tex
Vescovi's comments and changes

\Abstract{
The abstract of the article.
}

@@

Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (\textless5 sec) of maximal efforts \citep{mangineSpeedForcePower2014}, whereas the ability to achieve maximal sprint speed varies based on the type of sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60 m) \citep{ward-smithEnergyConversionStrategies2001}, but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40 m)\citep{brownAssessmentLinearSprinting2004}.

Describing individual differences in short sprint performance, quantifying training intervention effects, but more importantly, understanding short sprint \emph{bottle necks} (i.e.~limiting factors) has been one of the prominent tasks of sports science profession since its establishment. The \pkg{shorts} package \citep{R-shorts}, written in the R language \citep{R-base}, represents an open-source tool that aims to help sports scientist in achieving these goals through mathematical modeling.

In the current paper, we will provide explanation of one commonly used mathematical equation to model short sprints, modeling applications using the \pkg{shorts} package, issues that can arise during measurement and estimation, and potential solutions to those problems.

@@

The Figure \ref{fig:four-athletes-kinematics} depicts distance, velocity, and acceleration over time (from 0 to 6s).

@@

If we plot acceleration against velocity (Figure \ref{fig:four-athletes-profile}), we will get Acceleration-Velocity profile, which is linear, according to the mathematical model.

@@

Short sprints profiling is usually performed by: (1) measuring split times using timing gates (i.e., positioned at various distances, e.g., 5, 10, 20, 30, 40m), and (2) getting a velocity trace using a radar gun. Estimation of MSS and TAU parameters from equation \eqref{eq:velocity-time} is performed in \pkg{shorts} package using non-linear least squares regression implemented in the \texttt{nls} function in the \pkg{base} R \citep{R-base} and \texttt{nlme} function in the \pkg{nlme} package \citep{R-nlme} for the mixed-effect models.

@@

Let's consider an example of an athlete with MSS equal to 9 \(ms^-1\), TAU equal to 1.3, and MAC equal to 6.92 \(ms^-2\) performing 40m sprint with timing gates positioned at each 10m split. When it comes to split times, distance is a predictor, and time is outcome variable, thus the equation \eqref{eq:velocity-time} takes another form:

@@

\begin{CodeChunk}
\begin{CodeInput}
R> require(shorts)
R> 
R> split_distance <- c(10, 20, 30, 40)
R> 
R> split_time <- c(2.17, 3.43, 4.60, 5.73)
R> 
R> m1 <- shorts::model_using_splits(
+   distance = split_distance,
+   time = split_time
+ )
R> 
R> m1
\end{CodeInput}
\begin{CodeOutput}
#> Estimated model parameters
#> --------------------------
#>                 MSS                 TAU                 MAC 
#>                9.01                1.31                6.89 
#>                PMAX     time_correction distance_correction 
#>               15.52                0.00                0.00 
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE 
#>   0.00249   1.00000  -0.00178   0.00265   0.00265   0.00176 
#>       MAE      MAPE 
#>   0.00157   0.04742
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> summary(m1$model)
\end{CodeInput}
\begin{CodeOutput}
#> 
#> Formula: corrected_time ~ TAU * I(LambertW::W(-exp(1)^(-distance/(MSS * 
#>     TAU) - 1))) + distance/MSS + TAU
#> 
#> Parameters:
#>     Estimate Std. Error t value Pr(>|t|)    
#> MSS   9.0121     0.0155     581  3.0e-06 ***
#> TAU   1.3083     0.0066     198  2.5e-05 ***
#> ---
#> Signif. codes:  
#> 0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> Residual standard error: 0.00249 on 2 degrees of freedom
#> 
#> Number of iterations to convergence: 4 
#> Achieved convergence tolerance: 3.11e-06
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> df <- shorts::predict_kinematics(
+   m1,
+   max_time = 6,
+   frequency = 100,
+   # Additional parameters forwarded to shorts::get_air_resistance
+   # Otherwise, defaults are used
+   bodymass = 80,
+   bodyheight = 1.85,
+   barometric_pressure = 780,
+   air_temperature = 20,
+   wind_velocity = 0.5
+ )
R> 
R> head(df)
\end{CodeInput}
\begin{CodeOutput}
#>   time distance velocity acceleration air_resistance force
#> 1 0.00 0.000000   0.0000         6.89        0.07536   551
#> 2 0.01 0.000344   0.0686         6.84        0.05609   547
#> 3 0.02 0.001371   0.1367         6.78        0.03978   543
#> 4 0.03 0.003076   0.2043         6.73        0.02636   539
#> 5 0.04 0.005455   0.2714         6.68        0.01576   534
#> 6 0.05 0.008502   0.3379         6.63        0.00792   530
#>   power relative_power
#> 1   0.0          0.000
#> 2  37.5          0.469
#> 3  74.2          0.928
#> 4 110.0          1.375
#> 5 145.0          1.813
#> 6 179.2          2.241
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> data(split_times)
R> 
R> head(split_times)
\end{CodeInput}
\begin{CodeOutput}
#> # A tibble: 6 x 4
#>   athlete bodyweight distance     time
#>   <chr>        <dbl>    <dbl> <I<dbl>>
#> 1 John            75        5     1.20
#> 2 John            75       10     1.97
#> 3 John            75       15     2.66
#> 4 John            75       20     3.31
#> 5 John            75       30     4.59
#> 6 John            75       40     5.85
\end{CodeOutput}
\begin{CodeInput}
R> 
R> # Mixed model
R> m2 <- shorts::mixed_model_using_splits(
+   data = split_times,
+   distance = "distance",
+   time = "time",
+   athlete = "athlete",
+ 
+   # Select random effects
+   # Default is MSS and TAU
+   random = MSS + TAU ~ 1
+ )
R> 
R> m2
\end{CodeInput}
\begin{CodeOutput}
#> Estimated fixed model parameters
#> --------------------------------
#>                 MSS                 TAU                 MAC 
#>               8.065               0.655              12.309 
#>                PMAX     time_correction distance_correction 
#>              24.818               0.000               0.000 
#> 
#> Estimated random model parameters
#> ----------------------------------
#>     athlete  MSS   TAU  MAC PMAX time_correction
#> 1     James 9.69 0.847 11.4 27.7               0
#> 2       Jim 7.83 0.505 15.5 30.4               0
#> 3      John 7.78 0.727 10.7 20.8               0
#> 4 Kimberley 8.57 0.802 10.7 22.9               0
#> 5  Samantha 6.45 0.395 16.3 26.4               0
#>   distance_correction
#> 1                   0
#> 2                   0
#> 3                   0
#> 4                   0
#> 5                   0
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE 
#>    0.0260    0.9998   -0.0293    0.0496    0.0496    0.0214 
#>       MAE      MAPE 
#>    0.0172    0.9019
\end{CodeOutput}
\end{CodeChunk}

@@

Estimation of the short sprint profile using radar gun data takes time as predictor and velocity as an target or outcome variable. Thus equation \eqref{eq:velocity-time} is used to estimate MSS and TAU.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> sprint_time <- seq(0, 6, 1)
R> 
R> sprint_velocity <- c(0.00, 4.83, 7.07, 8.10, 8.59, 8.81, 8.91)
R> 
R> m3 <- shorts::model_using_radar(
+   velocity = sprint_velocity,
+   time = sprint_time
+ )
R> 
R> m3
\end{CodeInput}
\begin{CodeOutput}
#> Estimated model parameters
#> --------------------------
#>                 MSS                 TAU                 MAC 
#>                9.00                1.30                6.92 
#>                PMAX     time_correction distance_correction 
#>               15.58                0.00                0.00 
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE 
#>   0.00327   1.00000  -0.00406   0.00532   0.00532   0.00276 
#>       MAE      MAPE 
#>   0.00207       NaN
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> m3_weighted <- shorts::model_using_radar(
+   velocity = sprint_velocity,
+   time = sprint_time,
+   weights = 1 / (sprint_velocity + 1)
+ )
R> 
R> m3_weighted
\end{CodeInput}
\begin{CodeOutput}
#> Estimated model parameters
#> --------------------------
#>                 MSS                 TAU                 MAC 
#>                9.00                1.30                6.92 
#>                PMAX     time_correction distance_correction 
#>               15.58                0.00                0.00 
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE 
#>   0.00108   1.00000  -0.00406   0.00534   0.00534   0.00276 
#>       MAE      MAPE 
#>   0.00206       NaN
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> data("radar_gun_data")
R> 
R> head(radar_gun_data)
\end{CodeInput}
\begin{CodeOutput}
#> # A tibble: 6 x 4
#>   athlete bodyweight  time velocity
#>   <chr>        <dbl> <dbl>    <dbl>
#> 1 John            75  0       0    
#> 2 John            75  0.01    0.075
#> 3 John            75  0.02    0.149
#> 4 John            75  0.03    0.222
#> 5 John            75  0.04    0.291
#> 6 John            75  0.05    0.367
\end{CodeOutput}
\begin{CodeInput}
R> 
R> m4 <- shorts::mixed_model_using_radar(
+   radar_gun_data,
+   time = "time",
+   velocity = "velocity",
+   athlete = "athlete"
+ )
R> 
R> m4
\end{CodeInput}
\begin{CodeOutput}
#> Estimated fixed model parameters
#> --------------------------------
#>                 MSS                 TAU                 MAC 
#>                8.30                1.01                8.24 
#>                PMAX     time_correction distance_correction 
#>               17.09                0.00                0.00 
#> 
#> Estimated random model parameters
#> ----------------------------------
#>     athlete   MSS   TAU  MAC PMAX time_correction
#> 1     James 10.00 1.111 9.00 22.5               0
#> 2       Jim  8.00 0.889 9.00 18.0               0
#> 3      John  8.00 1.069 7.48 15.0               0
#> 4 Kimberley  9.01 1.286 7.01 15.8               0
#> 5  Samantha  6.50 0.685 9.50 15.4               0
#>   distance_correction
#> 1                   0
#> 2                   0
#> 3                   0
#> 4                   0
#> 5                   0
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE 
#>    0.0516    0.9994   -0.2191    0.1983    0.2191    0.0516 
#>       MAE      MAPE 
#>    0.0395       NaN
\end{CodeOutput}
\end{CodeChunk}

@@

There is a challenge when collecting sprint data that could have a substantial impact on modeled outcomes. To ensure accurate parameter outcomes, the initial force production must be synced with start time. Below we describe this challenge when using radar guns or timing gates and suggest potential solutions within the \pkg{shorts} package.

@@

One source of error in the modeled estimation using radar gun is the time synchronization. In theory, synchronization is ideal when sprint is initiated at \(t=0\) (i.e., \(v(t=0) = 0\)). In practice, this is often not the case. Let's use our athlete and add and deduct 0.5second to simulate an error in synchronization and its effect on estimated MSS and TAU.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # Using the true time
R> predict_velocity_at_time(
+   time = df$`true time`,
+   MSS = m5$parameters$MSS,
+   TAU = m5$parameters$TAU
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 0.00 4.83 7.07 8.11 8.59 8.81 8.91
\end{CodeOutput}
\begin{CodeInput}
R> 
R> # Using time with sync issues
R> predict_velocity_at_time(
+   time = df$`0.5s added`,
+   MSS = m8$parameters$MSS,
+   TAU = m8$parameters$TAU,
+   time_correction = m8$parameters$time_correction
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 7.82e-05 4.83e+00 7.07e+00 8.11e+00 8.59e+00 8.81e+00
#> [7] 8.91e+00
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # Adding 0.5s to radar_gun_data
R> radar_gun_data$time <- radar_gun_data$time + 0.5
R> 
R> # Mixed model with time correction being fixed effect
R> m10 <- shorts::mixed_model_using_radar_with_time_correction(
+   radar_gun_data,
+   time = "time",
+   velocity = "velocity",
+   athlete = "athlete",
+   random = MSS + TAU ~ 1
+ )
R> 
R> m10
\end{CodeInput}
\begin{CodeOutput}
#> Estimated fixed model parameters
#> --------------------------------
#>                 MSS                 TAU                 MAC 
#>                8.30                1.01                8.24 
#>                PMAX     time_correction distance_correction 
#>               17.10               -0.50                0.00 
#> 
#> Estimated random model parameters
#> ----------------------------------
#>     athlete   MSS   TAU  MAC PMAX time_correction
#> 1     James 10.00 1.111 9.00 22.5            -0.5
#> 2       Jim  8.00 0.889 9.00 18.0            -0.5
#> 3      John  8.00 1.069 7.48 15.0            -0.5
#> 4 Kimberley  9.01 1.285 7.01 15.8            -0.5
#> 5  Samantha  6.50 0.685 9.50 15.4            -0.5
#>   distance_correction
#> 1                   0
#> 2                   0
#> 3                   0
#> 4                   0
#> 5                   0
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE 
#>    0.0516    0.9994   -0.2190    0.1983    0.2190    0.0516 
#>       MAE      MAPE 
#>    0.0395       Inf
\end{CodeOutput}
\begin{CodeInput}
R> 
R> 
R> # Mixed model with time correction being random effect
R> m11 <- shorts::mixed_model_using_radar_with_time_correction(
+   radar_gun_data,
+   time = "time",
+   velocity = "velocity",
+   athlete = "athlete",
+   random = MSS + TAU + time_correction ~ 1
+ )
R> 
R> m11
\end{CodeInput}
\begin{CodeOutput}
#> Estimated fixed model parameters
#> --------------------------------
#>                 MSS                 TAU                 MAC 
#>                8.30                1.01                8.24 
#>                PMAX     time_correction distance_correction 
#>               17.10               -0.50                0.00 
#> 
#> Estimated random model parameters
#> ----------------------------------
#>     athlete   MSS   TAU  MAC PMAX time_correction
#> 1     James 10.00 1.110 9.00 22.5            -0.5
#> 2       Jim  8.00 0.889 9.00 18.0            -0.5
#> 3      John  8.00 1.069 7.48 15.0            -0.5
#> 4 Kimberley  9.01 1.285 7.01 15.8            -0.5
#> 5  Samantha  6.50 0.685 9.50 15.4            -0.5
#>   distance_correction
#> 1                   0
#> 2                   0
#> 3                   0
#> 4                   0
#> 5                   0
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE 
#>    0.0516    0.9994   -0.2188    0.1982    0.2188    0.0516 
#>       MAE      MAPE 
#>    0.0395       Inf
\end{CodeOutput}
\end{CodeChunk}

@@

Let's imagine we have two twin brothers with same short sprint characteristics: MSS equal to 9\(ms^-1\), TAU equal to 1.3, and MAC equal to 6.92\(ms^-2\). Let's call them John and Jack. The are both performing 40m sprint using timing gates set at 5, 10, 20, 30, and 40m. The initial timing gate at the start (i.e., \(d=0m\)) serves to activate the timing system (i.e., when they cross the beam).

John represents the \emph{theoretical model}, in which we assume that the initial force production and the timing initiation are perfectly synchronized. Jack, on the other hand, represents a \emph{practical model}, and decides to move slightly behind the initial timing gate (i.e.~for 0.5m) and use body rocking to initiate the sprint start. In other words, Jack is using a \emph{flying start}, a common scenario when testing field sports athletes. Let's see how their sprints differ.

>>>>

\Abstract{
Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints have been modeled using the mono-exponential equation that involves two parameters: (1) maximum sprinting speed (MSS) and (2) relative acceleration (TAU). Short sprint performance data comes from two sources: (1) split times using timing gates and (2) radar gun. In this paper, we demonstrate how these data can be modeled using the \pkg{shorts} package and how timing issues (i.e., synchronization issues with the radar gun, or using the flying start with the timing gates), which are common in practical measurement settings, can introduce error in parameter estimates. Further, we demonstrated how different model definitions implemented in \pkg{shorts} package can alleviate these errors.
}

@@

Short sprint performance is one of the most distinguishable and admired physical trait in sports. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (\textless5 sec) of maximal efforts \citep{mangineSpeedForcePower2014}, whereas the ability to achieve maximal sprint speed varies based on the type of sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60 m) \citep{ward-smithEnergyConversionStrategies2001}, but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40 m)\citep{brownAssessmentLinearSprinting2004}. Regardless of the differences in kinematics between athletes, evaluating short sprint performance is routinely included within a battery of fitness tests for a wide range of sports

The use of force plates is considered the gold standard for assessing mechanical properties of sprinting; however, there are logistical and financial challenges to capturing the profile of an entire sprint \citep{morinSimpleMethodComputing2019, samozinoSimpleMethodMeasuring2016}. Radar and laser technology are frequently used laboratory-grade methods \citep{buchheitMechanicalDeterminantsAcceleration2014, edwardsSprintAccelerationCharacteristics2020, jimenez-reyesRelationshipVerticalHorizontal2018, marcote-pequenoAssociationForceVelocity2019} but not normally accessible to practitioners working in sports. Undoubtedly, the most common method available and used to evaluate sprint performance are timing gates. Often multiple gates are positioned at varying distances to capture split times (e.g., 5, 10, 20 m), which can now be incorporated into the method for determining sprint mechanical properties \citep{morinSimpleMethodComputing2019, samozinoSimpleMethodMeasuring2016}. This approach presents an advantage to practitioners who can use the outcomes to describe individual differences, quantify the effects of training interventions, and better understanding the limiting factors of performance. The \pkg{shorts} package \citep{R-shorts}, written in the R language \citep{R-base}, represents an open-source tool to help sport scientists translate raw timing data into detailed mechanical outcomes through mathematical modeling \citep{morinSimpleMethodComputing2019, samozinoSimpleMethodMeasuring2016}.

In the current paper, we will provide an explanation of one commonly used mathematical equation to model short sprints, modeling applications using the \pkg{shorts} package, issues that can arise during measurement and estimation, and potential solutions to those problems.

@@

Figure \ref{fig:four-athletes-kinematics} depicts distance, velocity, and acceleration over time (from 0 to 6s).

@@

Plotting acceleration against velocity (Figure \ref{fig:four-athletes-profile}), we will get Acceleration-Velocity profile, which is linear, according to the mathematical model.

@@

Short sprints profiling is usually performed by: (1) measuring split times using timing gates (i.e., positioned at various distances, e.g., 5, 10, 20, 30, 40m), (2) getting a velocity trace using a radar gun. Estimation of MSS and TAU parameters from equation \eqref{eq:velocity-time} is performed in \pkg{shorts} package using non-linear least squares regression implemented in the \texttt{nls} function in the \pkg{base} R \citep{R-base} and \texttt{nlme} function in the \pkg{nlme} package \citep{R-nlme} for the mixed-effect models.

@@

Let's consider an example of an athlete with MSS equal to 9 \(ms^-1\), TAU equal to 1.3, and MAC equal to 6.92 \(ms^-2\) performing 40m sprint with timing gates positioned at each 10m split. For split times, distance is a predictor, and time is the outcome variable, thus the equation \eqref{eq:velocity-time} becomes:

@@

\begin{CodeChunk}
\begin{CodeInput}
R> require(shorts)
R> 
R> split_distance <- c(10, 20, 30, 40)
R> 
R> split_time <- c(2.17, 3.43, 4.60, 5.73)
R> 
R> m1 <- shorts::model_using_splits(
+   distance = split_distance,
+   time = split_time
+ )
R> 
R> m1
\end{CodeInput}
\begin{CodeOutput}
#> Estimated model parameters
#> --------------------------
#>                 MSS                 TAU                 MAC 
#>                9.01                1.31                6.89 
#>                PMAX     time_correction distance_correction 
#>               15.52                0.00                0.00 
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>   0.00249   1.00000  -0.00178   0.00265   0.00265   0.00176   0.00157 
#>      MAPE 
#>   0.04742
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> summary(m1$model)
\end{CodeInput}
\begin{CodeOutput}
#> 
#> Formula: corrected_time ~ TAU * I(LambertW::W(-exp(1)^(-distance/(MSS * 
#>     TAU) - 1))) + distance/MSS + TAU
#> 
#> Parameters:
#>     Estimate Std. Error t value Pr(>|t|)    
#> MSS   9.0121     0.0155     581  3.0e-06 ***
#> TAU   1.3083     0.0066     198  2.5e-05 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> Residual standard error: 0.00249 on 2 degrees of freedom
#> 
#> Number of iterations to convergence: 4 
#> Achieved convergence tolerance: 3.11e-06
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> df <- shorts::predict_kinematics(
+   m1,
+   max_time = 6,
+   frequency = 100,
+   # Additional parameters forwarded to shorts::get_air_resistance
+   # Otherwise, defaults are used
+   bodymass = 80,
+   bodyheight = 1.85,
+   barometric_pressure = 780,
+   air_temperature = 20,
+   wind_velocity = 0.5
+ )
R> 
R> head(df)
\end{CodeInput}
\begin{CodeOutput}
#>   time distance velocity acceleration air_resistance force power
#> 1 0.00 0.000000   0.0000         6.89        0.07536   551   0.0
#> 2 0.01 0.000344   0.0686         6.84        0.05609   547  37.5
#> 3 0.02 0.001371   0.1367         6.78        0.03978   543  74.2
#> 4 0.03 0.003076   0.2043         6.73        0.02636   539 110.0
#> 5 0.04 0.005455   0.2714         6.68        0.01576   534 145.0
#> 6 0.05 0.008502   0.3379         6.63        0.00792   530 179.2
#>   relative_power
#> 1          0.000
#> 2          0.469
#> 3          0.928
#> 4          1.375
#> 5          1.813
#> 6          2.241
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> data(split_times)
R> 
R> head(split_times)
\end{CodeInput}
\begin{CodeOutput}
#> # A tibble: 6 x 4
#>   athlete bodyweight distance     time
#>   <chr>        <dbl>    <dbl> <I<dbl>>
#> 1 John            75        5     1.20
#> 2 John            75       10     1.97
#> 3 John            75       15     2.66
#> 4 John            75       20     3.31
#> 5 John            75       30     4.59
#> 6 John            75       40     5.85
\end{CodeOutput}
\begin{CodeInput}
R> 
R> # Mixed model
R> m2 <- shorts::mixed_model_using_splits(
+   data = split_times,
+   distance = "distance",
+   time = "time",
+   athlete = "athlete",
+ 
+   # Select random effects
+   # Default is MSS and TAU
+   random = MSS + TAU ~ 1
+ )
R> 
R> m2
\end{CodeInput}
\begin{CodeOutput}
#> Estimated fixed model parameters
#> --------------------------------
#>                 MSS                 TAU                 MAC 
#>               8.065               0.655              12.309 
#>                PMAX     time_correction distance_correction 
#>              24.818               0.000               0.000 
#> 
#> Estimated random model parameters
#> ----------------------------------
#>     athlete  MSS   TAU  MAC PMAX time_correction distance_correction
#> 1     James 9.69 0.847 11.4 27.7               0                   0
#> 2       Jim 7.83 0.505 15.5 30.4               0                   0
#> 3      John 7.78 0.727 10.7 20.8               0                   0
#> 4 Kimberley 8.57 0.802 10.7 22.9               0                   0
#> 5  Samantha 6.45 0.395 16.3 26.4               0                   0
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>    0.0260    0.9998   -0.0293    0.0496    0.0496    0.0214    0.0172 
#>      MAPE 
#>    0.9019
\end{CodeOutput}
\end{CodeChunk}

@@

Estimation of the short sprint profile using radar gun data takes time as predictor and velocity as the outcome variable. Thus equation \eqref{eq:velocity-time} is used to estimate MSS and TAU.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> sprint_time <- seq(0, 6, 1)
R> 
R> sprint_velocity <- c(0.00, 4.83, 7.07, 8.10, 8.59, 8.81, 8.91)
R> 
R> m3 <- shorts::model_using_radar(
+   velocity = sprint_velocity,
+   time = sprint_time
+ )
R> 
R> m3
\end{CodeInput}
\begin{CodeOutput}
#> Estimated model parameters
#> --------------------------
#>                 MSS                 TAU                 MAC 
#>                9.00                1.30                6.92 
#>                PMAX     time_correction distance_correction 
#>               15.58                0.00                0.00 
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>   0.00327   1.00000  -0.00406   0.00532   0.00532   0.00276   0.00207 
#>      MAPE 
#>       NaN
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> m3_weighted <- shorts::model_using_radar(
+   velocity = sprint_velocity,
+   time = sprint_time,
+   weights = 1 / (sprint_velocity + 1)
+ )
R> 
R> m3_weighted
\end{CodeInput}
\begin{CodeOutput}
#> Estimated model parameters
#> --------------------------
#>                 MSS                 TAU                 MAC 
#>                9.00                1.30                6.92 
#>                PMAX     time_correction distance_correction 
#>               15.58                0.00                0.00 
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>   0.00108   1.00000  -0.00406   0.00534   0.00534   0.00276   0.00206 
#>      MAPE 
#>       NaN
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> data("radar_gun_data")
R> 
R> head(radar_gun_data)
\end{CodeInput}
\begin{CodeOutput}
#> # A tibble: 6 x 4
#>   athlete bodyweight  time velocity
#>   <chr>        <dbl> <dbl>    <dbl>
#> 1 John            75  0       0    
#> 2 John            75  0.01    0.075
#> 3 John            75  0.02    0.149
#> 4 John            75  0.03    0.222
#> 5 John            75  0.04    0.291
#> 6 John            75  0.05    0.367
\end{CodeOutput}
\begin{CodeInput}
R> 
R> m4 <- shorts::mixed_model_using_radar(
+   radar_gun_data,
+   time = "time",
+   velocity = "velocity",
+   athlete = "athlete"
+ )
R> 
R> m4
\end{CodeInput}
\begin{CodeOutput}
#> Estimated fixed model parameters
#> --------------------------------
#>                 MSS                 TAU                 MAC 
#>                8.30                1.01                8.24 
#>                PMAX     time_correction distance_correction 
#>               17.09                0.00                0.00 
#> 
#> Estimated random model parameters
#> ----------------------------------
#>     athlete   MSS   TAU  MAC PMAX time_correction distance_correction
#> 1     James 10.00 1.111 9.00 22.5               0                   0
#> 2       Jim  8.00 0.889 9.00 18.0               0                   0
#> 3      John  8.00 1.069 7.48 15.0               0                   0
#> 4 Kimberley  9.01 1.286 7.01 15.8               0                   0
#> 5  Samantha  6.50 0.685 9.50 15.4               0                   0
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>    0.0516    0.9994   -0.2191    0.1983    0.2191    0.0516    0.0395 
#>      MAPE 
#>       NaN
\end{CodeOutput}
\end{CodeChunk}

@@

There is a challenge when collecting sprint data that could have a substantial impact on modeled outcomes. To ensure accurate parameter outcomes, the initial force production must be synced with start time \citep{haugenPowerForceVelocityProfilingSprinting2020, haugenSprintMechanicalVariables2019}. Below we describe this challenge when using radar guns or timing gates and suggest potential solutions within the \pkg{shorts} package.

@@

One source of error in the modeled estimation using a radar gun is the time synchronization. In theory, synchronization is ideal when a sprint is initiated at \(t=0\) (i.e., \(v(t=0) = 0\)). In practice, this is often not the case. Let's use our athlete and add and deduct 0.5s to simulate an error in synchronization and its effect on estimated MSS and TAU.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # Using the true time
R> predict_velocity_at_time(
+   time = df$`true time`,
+   MSS = m5$parameters$MSS,
+   TAU = m5$parameters$TAU
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 0.00 4.83 7.07 8.11 8.59 8.81 8.91
\end{CodeOutput}
\begin{CodeInput}
R> 
R> # Using time with sync issues
R> predict_velocity_at_time(
+   time = df$`0.5s added`,
+   MSS = m8$parameters$MSS,
+   TAU = m8$parameters$TAU,
+   time_correction = m8$parameters$time_correction
+ )
\end{CodeInput}
\begin{CodeOutput}
#> [1] 7.82e-05 4.83e+00 7.07e+00 8.11e+00 8.59e+00 8.81e+00 8.91e+00
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> # Adding 0.5s to radar_gun_data
R> radar_gun_data$time <- radar_gun_data$time + 0.5
R> 
R> # Mixed model with time correction being fixed effect
R> m10 <- shorts::mixed_model_using_radar_with_time_correction(
+   radar_gun_data,
+   time = "time",
+   velocity = "velocity",
+   athlete = "athlete",
+   random = MSS + TAU ~ 1
+ )
R> 
R> m10
\end{CodeInput}
\begin{CodeOutput}
#> Estimated fixed model parameters
#> --------------------------------
#>                 MSS                 TAU                 MAC 
#>                8.30                1.01                8.24 
#>                PMAX     time_correction distance_correction 
#>               17.10               -0.50                0.00 
#> 
#> Estimated random model parameters
#> ----------------------------------
#>     athlete   MSS   TAU  MAC PMAX time_correction distance_correction
#> 1     James 10.00 1.111 9.00 22.5            -0.5                   0
#> 2       Jim  8.00 0.889 9.00 18.0            -0.5                   0
#> 3      John  8.00 1.069 7.48 15.0            -0.5                   0
#> 4 Kimberley  9.01 1.285 7.01 15.8            -0.5                   0
#> 5  Samantha  6.50 0.685 9.50 15.4            -0.5                   0
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>    0.0516    0.9994   -0.2190    0.1983    0.2190    0.0516    0.0395 
#>      MAPE 
#>       Inf
\end{CodeOutput}
\begin{CodeInput}
R> 
R> 
R> # Mixed model with time correction being random effect
R> m11 <- shorts::mixed_model_using_radar_with_time_correction(
+   radar_gun_data,
+   time = "time",
+   velocity = "velocity",
+   athlete = "athlete",
+   random = MSS + TAU + time_correction ~ 1
+ )
R> 
R> m11
\end{CodeInput}
\begin{CodeOutput}
#> Estimated fixed model parameters
#> --------------------------------
#>                 MSS                 TAU                 MAC 
#>                8.30                1.01                8.24 
#>                PMAX     time_correction distance_correction 
#>               17.10               -0.50                0.00 
#> 
#> Estimated random model parameters
#> ----------------------------------
#>     athlete   MSS   TAU  MAC PMAX time_correction distance_correction
#> 1     James 10.00 1.110 9.00 22.5            -0.5                   0
#> 2       Jim  8.00 0.889 9.00 18.0            -0.5                   0
#> 3      John  8.00 1.069 7.48 15.0            -0.5                   0
#> 4 Kimberley  9.01 1.285 7.01 15.8            -0.5                   0
#> 5  Samantha  6.50 0.685 9.50 15.4            -0.5                   0
#> 
#> Model fit estimators
#> --------------------
#>       RSE R_squared    minErr    maxErr maxAbsErr      RMSE       MAE 
#>    0.0516    0.9994   -0.2188    0.1982    0.2188    0.0516    0.0395 
#>      MAPE 
#>       Inf
\end{CodeOutput}
\end{CodeChunk}

@@

Let's imagine we have two twin brothers with same short sprint characteristics: MSS equal to 9\(ms^-1\), TAU equal to 1.3, and MAC equal to 6.92\(ms^-2\). Let's call them John and Jack. They both perform 40m sprint using timing gates set at 5, 10, 20, 30, and 40m. The initial timing gate at the start (i.e., \(d=0m\)) serves to activate the timing system (i.e., when they cross the beam).

John represents the \emph{theoretical model}, in which we assume that the initial force production and the timing initiation are perfectly synchronized. Jack, on the other hand, represents a \emph{practical model}, and decides to move slightly behind the initial timing gate (i.e.~for 0.5m) and use body rocking to initiate the sprint start. In other words, Jack is using a \emph{flying start}, a common scenario when testing field sports athletes. Let's see how their sprint outcomes differ.
